Equivalence Hypergraphs: E-Graphs for Monoidal Theories

### 2023-10-31 sections/combinatorial_semantics.tex
Minor fixes

    % \update{Now $<_{c,{X+Y}} : V_{X} + V_{Y} + E_{X} + E_{Y} \to E_{X} + E_{Y} + \bot_{X} + \bot_{Y}$
    % And let's get back to the case when $<_{c,Z}^{\mu}(z) = \bot_{Z}$, $<_{c,X}^{\mu}(f_{V}(z)) = \bot_{X}$, $f_{E}(\bot_{Z}) = \bot_{X}$ and $g_{E}(\bot_{Z}) = <_{c,Y}^{\mu}(g_{V}(z))$
    % We then want to show that $[<_{c,X+Y}^{\mu}(f_{V};i_{1,V}(z))] = [<_{c,X+Y}^{\mu}(g_{V};i_{2,V}(z))]$
    % Note that we have $[i_{1,E}(\bot_{X})] = [i_{2,E}(<_{c,Y}^{\mu}(g_{V}(z)))]$ and $[<_{c,X+Y}^{\mu}(f_{V};i_{1,V}(z))] = [i_{1,V}(\bot_{X})]$ because $i_{1,V}$ is a coproduct injection and $<_{c}^{\mu}(\bot) = \bot$.
    % Now, because $g_{V};i_{2,V}$ is a homomorphism: $<_{c,X+Y}^{\mu}(g_{V};i_{2,V}(z)) = g_{V};i_{2,V}(<_{c,Z}(z)) = i_{2,V}(<_{c,Y}^{\mu}(g_{V}(z)))$
    % }
    We then quotient the sets of edges and nodes in $X + Y$ by $R_{V},R_{E}$
    \[
        \sfrac{X + Y}{\sim (R_{V},R_{E})} = \{\sfrac{V_{X} + V_{Y}}{\sim R_{V}}, \sfrac{E_{X} + E_{Y}}{\sim R_{E}}, \sfrac{s_{X+Y}}{\sim (R_{V},R_{E})}, \sfrac{t_{X,Y}}{\sim (R_{V},R_{E})}, \#_{\sfrac{X+Y}{\sim (R_{V},R_{E})}}, <_{c,\sfrac{X + Y}{\sim (R_{V},R_{E})}} \}    
    \]
    where quotienting on source and target functions is taken as in $Hyp_{\Sigma}$ and we will further omit writing the relations explicitly.
    Recall that $<_{c}$ is essentially a transitive clousure of $<_{c}^{\mu}$ and hence we can first define $<_{c,\sfrac{X+Y}{\sim}}^{\mu}([u]_{E})$ and $<_{c,\sfrac{X+Y}{\sim}}^{\mu}([u]_{V})$.
    For this we will consider several cases where \update{$f$ and $g$ are mono. I require mono just to simplify things, we can get rid of this requirement if everything else is ok.}.
    Assume that $u$ is an element of $E_{X+Y}$ (the case for a node is similar).
    \begin{enumerate}
        \item $u$ has a pre-image $z$ in $E_{Z}$, $[z) \not = \varnothing$.
              The last requirement means that there necessarily exist $x = <_{c,X}^{\mu}(f_{E}(z)) = f_{E}(<_{c,Z}^{\mu}(z))$ and $y = <_{c,Y}^{\mu}(g_{E}(z)) = g_{E}(<_{c,Z}(z))$.
              We then define $<_{c,\sfrac{X+Y}{\sim}}^{\mu}([u]) = [<_{c,X+Y}^{\mu}(u)]$
        \item $u$ has a pre-image $z$ in $E_{Z}$, $[z) = \varnothing$ and $[f(z)) = \varnothing$ $[g(z)) \not = \varnothing$.
              The last condition means that there exists $e_y = <_{c,Y}^{\mu}(g(z))$.
              We then define $<_{c,\sfrac{X+Y}{\sim}}^{\mu}([u]) = [g_{E};<_{c,Y}^{\mu};i_{2,E}(z)] = [i_{2,E}(e_y)]$.
        \item $u$ has a pre-image $z$ in $E_{Z}$, $[z) = \varnothing$ and $[f(z)) \not = \varnothing$ $[g(z)) = \varnothing$.
              The one before the last condition means that there exists $e_x = <_{c,X}^{\mu}(f(z))$.
              We then define $<_{c,\sfrac{X+Y}{\sim}}^{\mu}([u]) = [f_{E};<_{c,X}^{\mu};i_{1,E}(z)] = [i_{1,E}(e_x)]$.
        \item $u$ has no pre-image in $Z$ and then $<_{c,\sfrac{X+Y}{\sim}}^{\mu}([u]) = [<_{c,X+Y}(u)]$
    \end{enumerate}
    Clearly, all the cases above are disjoint.
    Let's check that this is a well-defined function.
    Suppose $t_1 \sim t_2$ then $<_{c,\sfrac{X+Y}{\sim}}^{\mu}([t_1]) = <_{c,\sfrac{X+Y}{\sim}}^{\mu}([t_2])$.
    $t_1 \sim t_2$ implies there is $z$ such that $t_1 = f_{E};i_{1,E}(z)$ and $t_2 = g_{E};i_{2,E}(z)$.
    Then we need to consider the first three cases outlined above.
    \begin{itemize}
        \item If there exists $z' = <_{c,Z}^{\mu}(z)$ then 
        \[
            [<_{c,X+Y}^{\mu}(f_{E};i_{1,E}(z))] = [f_{E};i_{1,E}(<_{c,Z}^{\mu}(z))] = [f_{E};i_{1,E}(z')]
        \]
        \[
            [<_{c,X+Y}^{\mu}(g_{E};i_{2,E}(z))] = [g_{E};i_{2,E}(<_{c,Z}^{\mu}(z))] = [g_{E};i_{2,E}(z')]
        \]
        and
        \[ 
            [f_{e};i_{1,E}(z')] = [g_{e};i_{2,E}(z')]
        \]
        by commutativity.
        \item The other conditions satisfy functionality by definition.
    \end{itemize}

    We also need to makes sure that $[]_{E} : E_{X} + E_{Y} \to \sfrac{E_{X} + E_{Y}}{\sim}$ and $[]_{V} : V_{X} + V_{Y} \to \sfrac{V_{X} + V_{Y}}{\sim}$ are homomorphisms.
    \begin{enumerate}
        \item $s_{X+Y};[]_{V}^{*}(e) = []_{E};s_{\sfrac{X+Y}{\sim}}(e)$
              This is the same as in $\catname{Hyp_{\Sigma}}$.
        \item $[<_{c,X+Y}^{\mu}(e)]_{E} = <_{c,\sfrac{X+Y}{\sim}}^{\mu}([e]_{E})$ for $e$ such that $[e) \not = \varnothing$.
              Let's make case analysis on $[e]_{E}$.
              \begin{itemize}
                \item $e$ has a pre-image $z$ in $E_{Z}$.
                    \begin{itemize}
                        \item If $[z) \not = \varnothing$.
                              Then, by definition, $<_{c,\sfrac{X+Y}{\sim}}^{\mu}([e]_{E}) = [<_{c,X+Y}(e)]_{E}$ which is what we need.
                        \item If $[z) = \varnothing$ and $[f_{E}(z)) = \varnothing$ and $[g_{E}(z)) \not = \varnothing$.
                              Because $i_{2,E}$ is a homomorphism, we have $<_{c,X+Y}^{\mu}(g_{E};i_{2,E}(z)) = i_{2,E}(<_{c,Y}(g_{E}(z))) = i_{2,E}(e_{y})$.
                              We next have either $e = f_{E};i_{1,E}(z)$ in which case $[e) = \varnothing$ because $i_{1,E}$ is a coproduct injection and \update{we do not require homomorphism-ness for such $e$}.
                              Or, $e = g_{E};i_{2,E}(z)$ and 
                              \[
                              <_{c,\sfrac{X+Y}{\sim}}([e]) = [g_{E};<_{c,Y}{\mu};i_{2,E}(z)]_{E} = [i_{2,E}(e_{y})]
                              \]
                              and
                              \begin{align*}
                                [<_{c,X+Y}(e)] = [<_{c,X+Y}(g_{E};i_{2,E}(z))] &= \\
                                &\text{because $[g_{E}(z)) \not = \varnothing$ and $i_{2,E}$ is homomorphism}\\
                                 = [i_{2,E}(<_{c,Y}(g_{E}(z)))] = [i_{2,E}(e_{y})]
                              \end{align*}
                              so we get
                              \[
                              <_{c,\sfrac{X+Y}{\sim}}([e]) = [<_{c,X+Y}(e)]
                              \]
                        \item The remaining case is analogous.
                    \end{itemize}
                \item If $e$ has no pre-image in $E_{Z}$ then homomorphism-ness of $[]_{E}$ is similar to the first case above.
              \end{itemize}
    \end{enumerate}
    The proof for when $e$ is a node is analogous.

>>>>

    % \update{Now $<_{c,{X+Y}} : V_{X} + V_{Y} + E_{X} + E_{Y} \to E_{X} + E_{Y} + \bot_{X} + \bot_{Y}$
    % And let's get back to the case when $<_{c,Z}^{\mu}(z) = \bot_{Z}$, $<_{c,X}^{\mu}(f_{V}(z)) = \bot_{X}$, $f_{E}(\bot_{Z}) = \bot_{X}$ and $g_{E}(\bot_{Z}) = <_{c,Y}^{\mu}(g_{V}(z))$
    % We then want to show that $[<_{c,X+Y}^{\mu}(f_{V};i_{1,V}(z))] = [<_{c,X+Y}^{\mu}(g_{V};i_{2,V}(z))]$
    % Note that we have $[i_{1,E}(\bot_{X})] = [i_{2,E}(<_{c,Y}^{\mu}(g_{V}(z)))]$ and $[<_{c,X+Y}^{\mu}(f_{V};i_{1,V}(z))] = [i_{1,V}(\bot_{X})]$ because $i_{1,V}$ is a coproduct injection and $<_{c}^{\mu}(\bot) = \bot$.
    % Now, because $g_{V};i_{2,V}$ is a homomorphism: $<_{c,X+Y}^{\mu}(g_{V};i_{2,V}(z)) = g_{V};i_{2,V}(<_{c,Z}(z)) = i_{2,V}(<_{c,Y}^{\mu}(g_{V}(z)))$
    % }
    We then quotient the sets of edges and nodes in $X + Y$ by $R_{V},R_{E}$
    \[
        \sfrac{X + Y}{\sim (R_{V},R_{E})} = \{\sfrac{V_{X} + V_{Y}}{\sim R_{V}}, \sfrac{E_{X} + E_{Y}}{\sim R_{E}}, \sfrac{s_{X+Y}}{\sim (R_{V},R_{E})}, \sfrac{t_{X,Y}}{\sim (R_{V},R_{E})}, \#_{\sfrac{X+Y}{\sim (R_{V},R_{E})}}, <_{c,\sfrac{X + Y}{\sim (R_{V},R_{E})}} \}    
    \]
    where quotienting on source and target functions is taken as in $Hyp_{\Sigma}$ and we will further omit writing the relations explicitly.
    Recall that $<_{c}$ is essentially a transitive clousure of $<_{c}^{\mu}$ and hence we can first define $<_{c,\sfrac{X+Y}{\sim}}^{\mu}([u]_{E})$ and $<_{c,\sfrac{X+Y}{\sim}}^{\mu}([u]_{V})$.
    For this we will consider several cases where \update{$f$ and $g$ are mono. I require mono just to simplify things, we can get rid of this requirement if everything else is ok.}.
    Assume that $u$ is an element of $E_{X+Y}$ (the case for a node is similar).
    \begin{enumerate}
        \item $u$ has a pre-image $z$ in $E_{Z}$, $[z) \not = \varnothing$.
              The last requirement means that there necessarily exist $x = <_{c,X}^{\mu}(f_{E}(z)) = f_{E}(<_{c,Z}^{\mu}(z))$ and $y = <_{c,Y}^{\mu}(g_{E}(z)) = g_{E}(<_{c,Z}^{\mu}(z))$.
              We then define $<_{c,\sfrac{X+Y}{\sim}}^{\mu}([u]) = [<_{c,X+Y}^{\mu}(u)]$
        \item $u$ has a pre-image $z$ in $E_{Z}$, $[z) = \varnothing$ and $[f(z)) = \varnothing$ $[g(z)) \not = \varnothing$.
              The last condition means that there exists $e_y = <_{c,Y}^{\mu}(g(z))$.
              We then define $<_{c,\sfrac{X+Y}{\sim}}^{\mu}([u]) = [g_{E};<_{c,Y}^{\mu};i_{2,E}(z)] = [i_{2,E}(e_y)]$.
        \item $u$ has a pre-image $z$ in $E_{Z}$, $[z) = \varnothing$ and $[f(z)) \not = \varnothing$ $[g(z)) = \varnothing$.
              The one before the last condition means that there exists $e_x = <_{c,X}^{\mu}(f(z))$.
              We then define $<_{c,\sfrac{X+Y}{\sim}}^{\mu}([u]) = [f_{E};<_{c,X}^{\mu};i_{1,E}(z)] = [i_{1,E}(e_x)]$.
        \item $u$ has no pre-image in $Z$ and then $<_{c,\sfrac{X+Y}{\sim}}^{\mu}([u]) = [<_{c,X+Y}^{\mu}(u)]$
    \end{enumerate}
    Clearly, all the cases above are disjoint.
    Let's check that this is a well-defined function.
    Suppose $t_1 \sim t_2$ then $<_{c,\sfrac{X+Y}{\sim}}^{\mu}([t_1]) = <_{c,\sfrac{X+Y}{\sim}}^{\mu}([t_2])$.
    $t_1 \sim t_2$ implies there is $z$ such that $t_1 = f_{E};i_{1,E}(z)$ and $t_2 = g_{E};i_{2,E}(z)$.
    Then we need to consider the first three cases outlined above.
    \begin{itemize}
        \item If there exists $z' = <_{c,Z}^{\mu}(z)$ then 
        \[
            [<_{c,X+Y}^{\mu}(f_{E};i_{1,E}(z))] = [f_{E};i_{1,E}(<_{c,Z}^{\mu}(z))] = [f_{E};i_{1,E}(z')]
        \]
        \[
            [<_{c,X+Y}^{\mu}(g_{E};i_{2,E}(z))] = [g_{E};i_{2,E}(<_{c,Z}^{\mu}(z))] = [g_{E};i_{2,E}(z')]
        \]
        and
        \[ 
            [f_{E};i_{1,E}(z')] = [g_{E};i_{2,E}(z')]
        \]
        by commutativity.
        \item The other conditions satisfy functionality by definition.
    \end{itemize}

    We also need to makes sure that $[]_{E} : E_{X} + E_{Y} \to \sfrac{E_{X} + E_{Y}}{\sim}$ and $[]_{V} : V_{X} + V_{Y} \to \sfrac{V_{X} + V_{Y}}{\sim}$ are homomorphisms.
    \begin{enumerate}
        \item $s_{X+Y};[]_{V}^{*}(e) = []_{E};s_{\sfrac{X+Y}{\sim}}(e)$
              This is the same as in $\catname{Hyp_{\Sigma}}$.
        \item $[<_{c,X+Y}^{\mu}(e)]_{E} = <_{c,\sfrac{X+Y}{\sim}}^{\mu}([e]_{E})$ for $e$ such that $[e) \not = \varnothing$.
              Let's make case analysis on $[e]_{E}$.
              \begin{itemize}
                \item $e$ has a pre-image $z$ in $E_{Z}$.
                    \begin{itemize}
                        \item If $[z) \not = \varnothing$.
                              Then, by definition, $<_{c,\sfrac{X+Y}{\sim}}^{\mu}([e]_{E}) = [<_{c,X+Y}^{\mu}(e)]_{E}$ which is what we need.
                        \item If $[z) = \varnothing$ and $[f_{E}(z)) = \varnothing$ and $[g_{E}(z)) \not = \varnothing$.
                              Because $i_{2,E}$ is a homomorphism, we have $<_{c,X+Y}^{\mu}(g_{E};i_{2,E}(z)) = i_{2,E}(<_{c,Y}^{\mu}(g_{E}(z))) = i_{2,E}(e_{y})$.
                              We next have either $e = f_{E};i_{1,E}(z)$ in which case $[e) = \varnothing$ because $i_{1,E}$ is a coproduct injection and \update{we do not require homomorphism-ness for such $e$}.
                              Or, $e = g_{E};i_{2,E}(z)$ and 
                              \[
                              <_{c,\sfrac{X+Y}{\sim}}^{\mu}([e]) = [g_{E};<_{c,Y}^{\mu};i_{2,E}(z)]_{E} = [i_{2,E}(e_{y})]
                              \]
                              and
                              \begin{align*}
                                [<_{c,X+Y}^{\mu}(e)] = [<_{c,X+Y}^{\mu}(g_{E};i_{2,E}(z))] &= \\
                                &\text{because $[g_{E}(z)) \not = \varnothing$ and $i_{2,E}$ is homomorphism}\\
                                 = [i_{2,E}(<_{c,Y}(g_{E}(z)))] = [i_{2,E}(e_{y})]
                              \end{align*}
                              so we get
                              \[
                              <_{c,\sfrac{X+Y}{\sim}}^{\mu}([e]) = [<_{c,X+Y}^{\mu}(e)]
                              \]
                        \item The remaining case is analogous.
                    \end{itemize}
                \item If $e$ has no pre-image in $E_{Z}$ then homomorphism-ness of $[]_{E}$ is similar to the first case above.
              \end{itemize}
    \end{enumerate}
    The proof for when $e$ is a node is analogous.
